R for Clinical Study Reports and Submission

### 2021-11-01 project-folder.Rmd
project-folder.Rmd edited online with Bitbucket
use sentence case instead of title case for headings; run styler on Rmds
fix grammar, trailing spaces, file organization

# (PART) Clinical Trial Project {-}

# Overview

In a late-stage clinical trial, the number of A&R deliverables can easily be in the hundreds. 
For an organization, it is common to have multiple ongoing clinical trials in a clinical program.

To deliver the A&R results of a clinical trial project, it is teamwork that typically requires collaborations from both statisticians and programmers. 
In this part, let's consider how to organize a clinical trial project as an A&R leader.

In Chapter \@ref(folder), we will discuss how to organize source code, documents, deliverables in an A&R clinical project. 
We recommend using the R package folder structure.

In Chapter \@ref(manage), we will discuss a process or system development lifecycle to manage the A&R of a clinical project. 
We recommend following an agile management approach to define, develop, validate and deliver work.

# Project Folder {#folder}

@@

- Consistency: everyone works on the same folder structure.
- Reproducibility: analysis can be executed and reproduced by different team members months/years later.  
- Automation: automatically check the integration of a project.  
- Compliance: reduce compliance issue

@@

For R users, you already benefit from a well-defined and consistent folder structure. 
That is the [R package folder structure](https://github.com/rstudio/cheatsheets/raw/master/package-development.pdf). 
Every R package developer is required to follow the same convention to organize their R functions before the R package can be disseminated through the Comprehensive R Archive Network (CRAN). 
As a user, you can easily install and use those R packages after installing from CRAN. 
There are many good resources to guide developers on R package development.
For example, Hadley Wickham's [R Packages book](https://r-pkgs.org/).

@@

In this book, we refer to an R package as **project-specific R package** 
if the purpose of an R package is to organize analysis scripts for a clinical project.

@@

- `*.Rproj`: RStudio project file used to open RStudio project.
- `DESCRIPTION`: Metadata for a package including authors, license, dependencies, etc.
- `R/`: Project-specific R functions.
- `vignettes/`: Analysis scripts using Rmarkdown.
- `man/`: Manual of project-specific R functions.

A general discussion of the R package folder structure can be found in Chapter 4 of the R package book (@wickham2015r).

@@

> Undeer the `vignettes/` folder, there are `adam/` folder and `tlf` folder.
> The `adam/` folder contains ADaM datasets. 
> The `tlf` folder contains output TLFs in RTF format.
> We put `adam/` and `tlf/` folders within the `vignettes` folder only for illustration purposes. 
> In an actual A&R report, you may have a different location to save your input and output.

@@

While creating those analysis scripts, we also defined a few helper functions (e.g. `fmt_num` and `count_by`). 
Those functions are saved in the `R/` folder.

```
R
├── count_by.R
└── fmt.R
```

With a formal project, it is also important to provide proper documentation for those help functions. 
We used Roxygen2 to document functions. 
For example, the header below defines each variable in `fmt_est`. 
More details can be found in [Chapter 10 of the R package book](https://r-pkgs.org/man.html)

@@

Those `Roxygen2` documentation will be transferred to 
standard R documentation `.Rd` files saved in the `man/` folder. 
This step is automatically handled by `devtools::document()` function.

@@

The `man` folder is used to save documentation automatically generated by `roxygen2`.
A typical workflow is to add `roxygen2` documentation at the header of each function in the `R/` folder. 
Then `devtools::documents()` is used to generate all the documentation files in the `man/` folder. 
More details can be found in [Chapter 10 of the R packages book](https://r-pkgs.org/man.html).

@@

Reproducibility of analysis is one of the most important aspects of regulatory deliverables. 
To ensure a successful reproduce, we need a controlled R environment, including the control of R version and control of R package version.  
By using the R package folder structure and proper tools (e.g. `renv`, `packrat`), 
we illustrate how to achieve reproducibility for R and R package versions.

@@

First, we introduce the control of R version. 
In the `esubdemo` project, a reproducible environment is created when you open the 
`esubdemo.Rproj` from RStudio. 
When we open the `esubdemo` project, RStudio will execute the R code in `.Rprofile` automatically.
So we can use `.Rprofile` to set up a reproducible environment. 
More details can be found in [https://rstats.wtf/r-startup.html](https://rstats.wtf/r-startup.html). 
After we open the `esubdemo` project, the code in `.Rprofile` will automatically check the current R version is the same as we defined in `.Rprofile`.
```{r, eval = FALSE}
# set up project R version
R_version <- "4.1.1"
```
If the R version mismatch, an error message is displayed as below.

@@

> `.Rprofile` is only for project-specific R package. 
> A standard R package should not use `.Rprofile`.

Next, we introduce the control of R package version, which is controlled in two layers. 
Firstly, we define a snapshot date in `.Rprofile`. 
The snapshot date allows us to freeze the source code repository.

@@

The R code above in the `.Rprofile` initiates the `renv` running environment.  
As a user, you can use `renv::init()`, `renv::snapshot()` and `renv::restore()` 
to initialize, save and restore R packages used for the current analysis project.

@@

> `renv.lock` file and `renv/` folder are only for project-specific R package.
> A standard R package should not use `renv`.

@@

```
Below R package path are searching in order to find installed R pacakges in this R session:
    "/home/zhanyilo/github-repo/esubdemo/renv/library/R-4.1/x86_64-pc-linux-gnu"
    "/rtmp/RtmpT3ljoY/renv-system-library"   
```

> A cloud-based R environment (e.g., RStudio Workbench)
> can enhance the reproducibility within an organization
> by using the same operating system, R version, and R package versions for an A&R project.
> More details can be found in [https://environments.rstudio.com/](https://environments.rstudio.com/).

> A container solution like [Docker](https://github.com/rstudio/r-docker) could further enhance 
> the reproducibility across an organization at the operating system level, but beyond the scope of this book.

@@

- [`devtools`](https://devtools.r-lib.org/): make package development easier. 
  + A good overview can be found in [Chapter 2 of the R packages](https://r-pkgs.org/whole-game.html) book  
  + `devttols::load_all()`: load all functions in `R/` folder and running environment. 
  + `devtools::document()`: automatically create documentation using `roxygen2`.
  + `devttols::check()`: automatically perform compliance check as an R package. 
  + `devtools::build_site()`: automatically run analysis scripts in batch and create a `pkgdown` website. 
- [`usethis`](https://usethis.r-lib.org/): automates repetitive tasks that arise during project setup and development.
- [`testthat`](https://testthat.r-lib.org/): streamline testing code.
  + A discussion of using the `testthat` for an A&R project can be found in (@madhutest).  
- [`pkgdown`](https://pkgdown.r-lib.org/): generate static HTML documentation website for an R package
  + It also allows you to run all analysis code in batch.

@@

As the consistent folder is defined, it also becomes easier to create specific tools 
that fit for the analysis and reporting purpose. Below are few potential tools that can be helpful:

@@

For an R package, there are already criteria to ensure R package integrity.  
More details can be found in [Chapter 19 of the R package book](https://r-pkgs.org/r-cmd-check.html).

>>>>

# Project folder {#folder}

@@

- Consistency: everyone works on the same folder structure.
- Reproducibility: analysis can be executed and reproduced by different team members months/years later.
- Automation: automatically check the integration of a project.
- Compliance: reduce compliance issues.

@@

For R users, you already benefit from a well-defined and consistent folder structure.
That is the [R package folder structure](https://github.com/rstudio/cheatsheets/raw/master/package-development.pdf).
Every R package developer is required to follow the same convention to organize
their R functions before the R package can be disseminated through the Comprehensive R Archive Network (CRAN).
As a user, you can easily install and use those R packages after installing from CRAN.
There are many good resources to guide developers on R package development,
for example, the [R Packages book](https://r-pkgs.org/) by Hadley Wickham.

@@

In this book, we refer to an R package as a **project-specific R package**
if the purpose of an R package is to organize analysis scripts for a clinical project.

@@

- `*.Rproj`: RStudio project file used to open RStudio project.
- `DESCRIPTION`: Metadata for a package including authors, license, dependencies, etc.
- `R/`: Project-specific R functions.
- `vignettes/`: Analysis scripts using R Markdown.
- `man/`: Manual of project-specific R functions.

A general discussion of the R package folder structure can be found
in Chapter 4 of the R Packages book (@wickham2015r).

@@

> Under the `vignettes/` folder, there are `adam/` folder and `tlf` folder.
> The `adam/` folder contains ADaM datasets.
> The `tlf` folder contains output TLFs in RTF format.
> We put `adam/` and `tlf/` folders within the `vignettes` folder only for illustration purposes.
> In an actual A&R report, you may have a different location to save your input and output.

@@

While creating those analysis scripts, we also defined a few helper functions (e.g., `fmt_num` and `count_by`).
Those functions are saved in the `R/` folder.

```
R/
├── count_by.R
└── fmt.R
```

With a formal project, it is also important to provide proper documentation for those help functions.
We used `roxygen2` to document functions.
For example, the header below defines each variable in `fmt_est`.
More details can be found in [Chapter 10 of the R Packages book](https://r-pkgs.org/man.html).

@@

The `roxygen2` documentation will be converted into
standard R documentation format, saved as `.Rd` files in the `man/` folder.
This step is automatically handled by `devtools::document()`.

@@

The `man/` folder is used to save documentation automatically generated by `roxygen2`.
A typical workflow is to add `roxygen2` documentation before each function in the `R/` folder.
Then `devtools::documents()` is used to generate all the documentation files in the `man/` folder.
More details can be found in [Chapter 10 of the R Packages book](https://r-pkgs.org/man.html).

@@

Reproducibility of analysis is one of the most important aspects of regulatory deliverables.
To ensure a successful reproduce, we need a controlled R environment,
including the control of the R version and the R package versions.
By using the R package folder structure and proper tools (e.g., `renv`, `packrat`),
we illustrate how to achieve reproducibility for R and R package versions.

@@

First, we introduce the control of the R version.
In the `esubdemo` project, a reproducible environment is created when you open
the `esubdemo.Rproj` from RStudio.
When we open the `esubdemo` project, RStudio will execute the R code in `.Rprofile` automatically.
So we can use `.Rprofile` to set up a reproducible environment.
More details can be found in <https://rstats.wtf/r-startup.html>.
After we open the `esubdemo` project, the code in `.Rprofile` will
automatically check the current R version is the same as we defined in `.Rprofile`.

```{r, eval = FALSE}
# Set project R version
R_version <- "4.1.1"
```

If the R version mismatch, an error message is displayed as below.

@@

> `.Rprofile` is only for project-specific R packages.
> A standard R package should not use `.Rprofile`.

Next, we introduce the control of the R package version, which is controlled in two layers.
Firstly, we define a snapshot date in `.Rprofile`.
The snapshot date allows us to freeze the source code repository.

@@

The R code above in the `.Rprofile` initiates the `renv` running environment.
As a user, you can use `renv::init()`, `renv::snapshot()`, and `renv::restore()`
to initialize, save and restore R packages used for the current analysis project.

@@

> The `renv.lock` file and `renv/` folder are only for project-specific R package.
> A standard R package should not use `renv`.

@@

```
Below R package path are searching in order to find installed R packages in this R session:
    "/home/zhanyilo/github-repo/esubdemo/renv/library/R-4.1/x86_64-pc-linux-gnu"
    "/rtmp/RtmpT3ljoY/renv-system-library"
```

> A cloud-based R environment (e.g., RStudio Workbench)
> can enhance the reproducibility within an organization
> by using the same operating system, R version, and R package versions for an A&R project.
> More details can be found at <https://environments.rstudio.com/>.

> A container solution like [Docker](https://github.com/rstudio/r-docker) could further enhance
> the reproducibility across an organization at the operating system level but beyond the scope of this book.

@@

- [`devtools`](https://devtools.r-lib.org/): make package development easier.
  + A good overview can be found in [Chapter 2 of the R Packages book](https://r-pkgs.org/whole-game.html).
  + `devttols::load_all()`: load all functions in `R/` folder and running environment.
  + `devtools::document()`: automatically create documentation using `roxygen2`.
  + `devttols::check()`: automatically perform compliance check as an R package.
  + `devtools::build_site()`: automatically run analysis scripts in batch and create a `pkgdown` website.
- [`usethis`](https://usethis.r-lib.org/): automates repetitive tasks that arise during project setup and development.
- [`testthat`](https://testthat.r-lib.org/): streamline testing code.
  + A discussion of using the `testthat` for an A&R project can be found in (@madhutest).
- [`pkgdown`](https://pkgdown.r-lib.org/): generate static HTML documentation website for an R package
  + It also allows you to run all analysis code in batch.

@@

As the consistent folder is defined, it also becomes easier to create specific tools
that fit the analysis and reporting purpose. Below are a few potential tools that can be helpful:

@@

For an R package, there are already criteria to ensure R package integrity.
More details can be found in [Chapter 19 of the R Packages book](https://r-pkgs.org/r-cmd-check.html).

### 2021-11-01 project-management.Rmd
project-management.Rmd edited online with Bitbucket
use sentence case instead of title case for headings; run styler on Rmds
fix grammar, trailing spaces, file organization

# Project Management {#manage}

## Setting up for Success

@@

Although many factors could decide if a project could execute efficiently,
we believe a few aspects are critical for the long-term success,
especially when managing clinical data analysis projects at scale.

### Work as a Team

As a general principle, all the team members involved in a project
should take basic training on project management and understand how to work as
a development team. 
@teamgeek provides some valuable tips on this topic.
As always, setting a clear goal and following a system development lifecycle
(SDLC) are essential.

### Design Clean Code Architecture

@@

### Set Capability Boundaries

@@

### Contribute to the Community

@@

SDLC for the A&R deliverables can be defined in four stages:

- planning: a planning stage to define the scope of a project.  
- development: a development stage to implement target deliverables. 
- validation: a validation stage to verify target deliverables. 
- operation: an operation stage to deliver work to stakeholders.

Importantly, we should not consider SDLC as a linear process. 
For example, if the study team identifies new requirement in a development or validation stage, 
the team should return to the planning stage to discuss and align the scope. 
An [agile project management](https://www.atlassian.com/agile/project-management) approach is suitable 
and recommended to an A&R clinical development project.
The goal is to embrace an iterative approach that continuously improves target deliverables based on
frequent stakeholder feedback.

@@

The project lead should also set up a project folder as discussed in Chapter \@ref(folder).
The project initiation can be simplified by creating an 
[RStudio project template](https://rstudio.github.io/rstudio-extensions/rstudio_project_templates.html).

@@

After a project is initiated, the study team start to develop TLFs based on 
pre-defined mock-up tables assigned to each team member.

The analysis code and relevant description can be saved in R markdown files 
in the `vignettes` folder.

The use of R markdown allows developers to assemble narrative text, code, and its
comments in one place to simplify documentation. 
It would be helpful to create a template and define a name convention for all TLFs deliverables. 
For example, we can use `tlf_` prefix in the filename to indicate that the R Markdown is to deliver TLFs. 
Multiple TLFs with similar designs can be included in one R Markdown file.

For example, in `esubdemo` project, we have 6 R Markdown files to create TLFs.

If there is any project-specific R functions that need to be developed, 
the R functions can be placed in the `R/` folder as discussed in Section \@ref(consistency).

@@

Validation is an crucial stage to ensure the deliverables are accurate and consistent. 
After the development stage is completed, the project team needs to validate the deliverables, including R Markdown
files for TLFs deliverables and project-specific R functions. 
The level of validation is determined at the define stage.

@@

To validate the content of a table, we can save the last datasets ready for table generation as a `.Rdata` file.
A validator can reproduce the TLF and compares it with the original result saved in `.Rdata` file. 
A test is passed when the results match. 
Customers can directly review the formatting of the TLFs by
customers by comparing them with the mock-up.

To validate a figure, we can use [snapshot testing](https://testthat.r-lib.org/articles/snapshotting.html) strategy.

@@

After completing the validation, the validator updates the status in a validation tracker. 
The project lead reviews the tracking sheet to make sure all required activities in the SDLC are completed and
the tracking sheet has been filled correctly. 
The deliverables are ready for customer review after all the validation steps are completed. 
Any changes to the output requested by customers are documented.

@@

After completion of development and required validation of all A&R deliverables, project lead runs compliance check for a project-specific R package similar to other R packages. 
`devtools::check()` is a convenient way to run compliance checks or `R CMD check`. 
`R CMD check` is an automated check of the contents in the R package for frequently encountered issues, before submission to CRAN. 
Since the project-specific R package is not submitted to CRAN, some checks can be customized and skipped in `devtools::check()`. 
The project lead should work with the study team to ensure all reported errors, warnings, and notes by `devtools::check()` are fixed.

The project lead can also use the R package `pkgdown` to build a complete website for a project-specific R package. 
The `pkgdown` website is a convenient way to run all analyses in batch and 
integrate outputs in a website, which comprehensively covers project-specific R functions, TLF generation programs, outputs and validation tracking information, etc... 
For example, in the `esubdemo` project, we created the `pkgdown` website at <https://elong0527.github.io/esubdemo/>.

Many of the tasks in SDLC can be completed automatically. 
An organization can leverage CI/CD workflow to automatically enable those tasks 
such as running testing cases and creating a `pkgdown` website. 
For example, in `esubdemo` project, we setup [GitHub Actions](https://github.com/elong0527/esubdemo/actions)
for these work.
It can be setup by using `usethis::use_github_action()` function.

>>>>

# Project management {#manage}

## Setting up for success

@@

Although many factors could decide if a project could execute efficiently,
we believe a few aspects are critical for long-term success,
especially when managing clinical data analysis projects at scale.

### Work as a team

As a general principle, all the team members involved in a project
should take basic training on project management and understand how to work as
a development team.
@teamgeek provides some valuable tips on this topic.
As always, setting a clear goal and following a system development lifecycle
(SDLC) is essential.

### Design clean code architecture

@@

### Set capability boundaries

@@

### Contribute to the community

@@

SDLC for the A&R deliverables can be defined in four stages.

- Planning: a planning stage to define the scope of a project.
- Development: a development stage to implement target deliverables.
- Validation: a validation stage to verify target deliverables.
- Operation: an operation stage to deliver work to stakeholders.

Importantly, we should not consider SDLC as a linear process.
For example, if the study team identifies a new requirement in a development or validation stage,
the team should return to the planning stage to discuss and align the scope.
An [agile project management](https://www.atlassian.com/agile/project-management) approach is suitable
and recommended for an A&R clinical development project.
The goal is to embrace an iterative approach that continuously improves target deliverables based on
frequent stakeholder feedback.

@@

The project lead should also set up a project folder, as discussed in Chapter \@ref(folder).
The project initiation can be simplified by creating an
[RStudio project template](https://rstudio.github.io/rstudio-extensions/rstudio_project_templates.html).

@@

After a project is initiated, the study team starts to develop TLFs based on
pre-defined mock-up tables assigned to each team member.

The analysis code and relevant description can be saved in R Markdown files
in the `vignettes` folder.

The use of R Markdown allows developers to assemble narrative text, code, and its
comments in one place to simplify documentation.
It would be helpful to create a template and define a name convention for all TLFs deliverables.
For example, we can use the `tlf_` prefix in the filename to indicate that the R Markdown file is for delivering TLFs.
Multiple TLFs with similar designs can be included in one R Markdown file.

For example, in the `esubdemo` project, we have 6 R Markdown files to create TLFs.

If there are any project-specific R functions that need to be developed,
the R functions can be placed in the `R/` folder as discussed in Section \@ref(consistency).

@@

Validation is a crucial stage to ensure the deliverables are accurate and consistent.
After the development stage is completed, the project team needs to validate the deliverables, including R Markdown
files for TLFs deliverables and project-specific R functions.
The level of validation is determined at the define stage.

@@

To validate the content of a table, we can save the last datasets ready for table generation as a `.Rdata` file.
A validator can reproduce the TLF and compare it with the original result saved in the `.Rdata` file.
A test is passed when the results match.
Customers can directly review the formatting of the TLFs by comparing them with the mock-up.

To validate a figure, we can use the [snapshot testing](https://testthat.r-lib.org/articles/snapshotting.html) strategy.

@@

After completing the validation, the validator updates the status in a validation tracker.
The project lead reviews the tracking sheet to make sure all required activities in the SDLC are completed,
and the tracking sheet has been filled correctly.
The deliverables are ready for customer review after all the validation steps are completed.
Any changes to the output requested by customers are documented.

@@

After completion of development and required validation of all A&R deliverables,
the project lead runs compliance checks for a project-specific R package similar to other R packages.
`devtools::check()` is a convenient way to run compliance checks or `R CMD check`.
`R CMD check` is an automated check of the contents in the R package
for frequently encountered issues before submission to CRAN.
Since the project-specific R package is not submitted to CRAN, some checks can be customized and skipped in `devtools::check()`.
The project lead should work with the study team to ensure
all reported errors, warnings, and notes by `devtools::check()` are fixed.

The project lead can also use the R package `pkgdown` to build a complete website for a project-specific R package.
The `pkgdown` website is a convenient way to run all analyses in batch and
integrate outputs in a website, which comprehensively covers
project-specific R functions, TLF generation programs, outputs and validation tracking information, etc.
For example, in the `esubdemo` project, we created the `pkgdown` website at <https://elong0527.github.io/esubdemo/>.

Many of the tasks in SDLC can be completed automatically.
An organization can leverage CI/CD workflow to automatically enable those tasks,
such as running testing cases and creating a `pkgdown` website.
For example, in the `esubdemo` project, we set up
[GitHub Actions](https://github.com/elong0527/esubdemo/actions)
for it. This can be done by using `usethis::use_github_action()`.

### 2021-11-01 submission-package.Rmd
fix grammar, trailing spaces, file organization

In this chapter, we will first give a high-level overview of what
assets in the eCTD submission package we should focus on when
submitting R code.
Secondly, we will discuss how to prepare the proprietary R packages
(if any), and make them part of the submission package.
Lastly, we will provide reusable templates for updating ADaM Reviewer's Guide (ADRG) and Analysis Results Metadata (ARM),
so that the reviewers receive proper instructions to reproduce the
analysis results.

@@

The demo project (R package) we will prepare for submission is called [`esubdemo`](https://github.com/elong0527/esubdemo) available on GitHub.
You can download or clone it:

@@

R package dependencies is another major factor to consider
before preparing your proprietary R package for submission.

@@

To prepare R packages for submission, one needs to pack the packages into text files, and then verify if the file only contains ASSCII characters.
With packed packages, one can unpack and install them from the text file, too.

@@

Let's pack the `esubdemo` package into a text file.
Assume the source package path is `esubdemo/`,
you should be able to pack the package with a single pipe:

@@

What happened in the pipe?
The function `pkglite::collate()` evaluates a specified scope of
folders and files defined by a list of **file specifications**,
and generate a **file collection** object.
This file collection contains the metadata required to properly convert
the files into text which is then used by `pkglite::pack()`.
With this flow, you can define the scope of the R source package
to be packed for submission in a flexible yet principled way.

>>>>

In this chapter, we will first give a high-level overview of what
assets in the eCTD submission package we should focus on when
submitting R code.
Secondly, we will discuss how to prepare the proprietary R packages
(if any), and make them part of the submission package.
Lastly, we will provide reusable templates for updating
ADaM Reviewer's Guide (ADRG) and Analysis Results Metadata (ARM)
so that the reviewers receive proper instructions to reproduce the
analysis results.

@@

The demo project (R package) we will prepare for submission is called
[`esubdemo`](https://github.com/elong0527/esubdemo), available on GitHub.
You can download or clone it:

@@

R package dependency is another major factor to consider
before preparing your proprietary R package for submission.

@@

To prepare R packages for submission, one needs to pack the packages into text files, and then verify if the file only contains ASCII characters.
With packed packages, one can unpack and install them from the text file, too.

@@

Let's pack the `esubdemo` package into a text file.
Assume the source package path is `esubdemo/`.
You should be able to pack the package with a single pipe:

@@

What happened in the pipe?
The function `pkglite::collate()` evaluates a specified scope of
folders and files defined by a list of **file specifications**,
and generates a **file collection** object.
This file collection contains the metadata required to properly convert
the files into text which is then used by `pkglite::pack()`.
With this flow, you can define the scope of the R source package
to be packed for submission in a flexible yet principled way.

### 2021-11-01 tlf-baseline.Rmd
fix grammar, trailing spaces, file organization

As in previous chapters, we first read `adsl` dataset that contain all 
required information for baseline characteristics table.

@@

- Step 1: Read the data set.
- Step 2: Use `table1::table1()` to get the baseline characteristics table.
- Step 3: Transfer the output in Step 2 into a data frame that only contains ASCII character.
- Step 4: Define the format of the rtf table by using R package `r2rtf`.

>>>>

As in previous chapters, we first read the `adsl` dataset that contains all
the required information for the baseline characteristics table.

@@

- Step 1: Read the data set.
- Step 2: Use `table1::table1()` to get the baseline characteristics table.
- Step 3: Transfer the output in Step 2 into a data frame that only contains ASCII characters.
- Step 4: Define the format of the RTF table by using the R package `r2rtf`.

### 2021-11-01 tlf-disposition.Rmd
fix grammar, trailing spaces, file organization

The first step is to read in the relevant datasets into R.
For a disposition table, all the required information is saved in a Subject-level Analysis Dataset (ADSL).
The dataset is provided in `sas7bdat` format that is a SAS data format currently used in many clinical trial analysis and reporting.
The `haven` package is able to read in the dataset, while maintaining its attributes (e.g. variable labels).

@@

In the code below, we calculate the number and percentage of participants who discontinued the study for different reasons.
For display purposes, `paste0("    ", row)` is used to add leading spaces to mimick identation in the final report.

@@

- Step 1: Read subject level data into R, i.e. `adsl`.
- Step 2: Count participants in the analysis population and name the dataset `n_rand`.
- Step 3: Count the number and percentage of participants that discontinued the study, and name the dataset `n_disc`.
- Step 4: Count the number and percentage of participants that discontinued the study for different reasons, 
  and name the dataset `n_reason`.
- Step 5: Count the number and percentage of participants that completed the study, and name the dataset `n_complete`.
- Step 6: Bind `n_rand`, `n_disc`, `n_reason` and `n_complete` by row. 
- Step 7: Write the final table to RTF

>>>>

The first step is to read in the relevant datasets into R.
For a disposition table, all the required information is saved in a Subject-level Analysis Dataset (ADSL).
The dataset is provided in `sas7bdat` format that is a SAS data format currently used in many clinical trial analysis and reporting.
The `haven` package is able to read in the dataset, while maintaining its attributes (e.g., variable labels).

@@

In the code below, we calculate the number and percentage of participants who discontinued the study for different reasons.
For display purposes, `paste0("    ", row)` is used to add leading spaces to mimick indentation in the final report.

@@

- Step 1: Read subject level data into R, i.e., `adsl`.
- Step 2: Count participants in the analysis population and name the dataset `n_rand`.
- Step 3: Count the number and percentage of participants that discontinued the study, and name the dataset `n_disc`.
- Step 4: Count the number and percentage of participants that discontinued the study for different reasons,
  and name the dataset `n_reason`.
- Step 5: Count the number and percentage of participants that completed the study, and name the dataset `n_complete`.
- Step 6: Bind `n_rand`, `n_disc`, `n_reason`, and `n_complete` by row.
- Step 7: Write the final table to RTF

### 2021-11-01 tlf-efficacy.Rmd
fix grammar, trailing spaces, file organization

For efficacy analysis, we only analyze change from baseline glucose data at Week 24.

@@

To prepare analysis dataset, 
we need both `adsl` and `adlbc` datasets for this analysis.

@@

```{r}
ana %>%
  select(USUBJID, TRTPN, AVISIT, AVAL, BASE, CHG) %>%
  head(4)
```
## Helper functions

To prepare report dataframe, we created a few helper functions 
by using the `fmt_num` function defined in Chapter \@ref(population).

@@

We start to analyze the imputed data using ANCOVA model with treatment and baseline glucose as covariates.

@@

Then we use `r2rtf` to prepare the table format for `t1`.
We also highlight how to handle special character in this example.

Special characters `^` and `_` are used to define superscript and subscript of text. And `{}` is to define the part that will be impacted. 
For example, `{^a}` provide a superscript `a` for footnote notation.   
`r2rtf` also support most LaTeX characters. 
Examples can be found in
[`r2rtf` Get Started Page](https://merck.github.io/r2rtf/articles/r2rtf.html#special-character). 
The `text_convert` argument in `r2rtf_xxx` functions control whether convert special characters.

@@

- Step 1: Read the data into R, i.e., `adsl` and `adlb`.
- Step 2: Define the analysis dataset. In this example, we define two analysis datasets. The first dataset is the efficacy population (`gluc`). The second dataset is the collection of all records post baseline and on or before week 24 (`ana`).
- Step 3: Impute the missing values. In our example, we name the `ana` dataset after imputation as `ana_locf`.
- Step 4: Summarize `gluc`, i.e., calculate the mean and standard derivation, and then format it into a rtf table.
- Step 5: Summarize `ana_locf`, i.e., calculate the pairwise comparison by ANCOVA model, and then format it into a rtf table.
- Step 6: Rowly bind the output in Step 4 and Step 5.

>>>>

For efficacy analysis, we only analyze the change from baseline glucose data at week 24.

@@

To prepare the analysis dataset,
we need both `adsl` and `adlbc` datasets for this analysis.

@@

```{r}
ana %>%
  select(USUBJID, TRTPN, AVISIT, AVAL, BASE, CHG) %>%
  head(4)
```

## Helper functions

To prepare the report data frame, we created a few helper functions
by using the `fmt_num()` function defined in Chapter \@ref(population).

@@

We start to analyze the imputed data using the ANCOVA model with treatment and baseline glucose as covariates.

@@

Then we use `r2rtf` to prepare the table format for `t1`.
We also highlight how to handle special characters in this example.

Special characters `^` and `_` are used to define superscript and subscript of text. And `{}` is to define the part that will be impacted.
For example, `{^a}` provide a superscript `a` for footnote notation.
`r2rtf` also supports most LaTeX characters.
Examples can be found on the
[`r2rtf` get started page](https://merck.github.io/r2rtf/articles/r2rtf.html#special-character).
The `text_convert` argument in `r2rtf_*()` functions controls whether convert special characters.

@@

- Step 1: Read the data into R, i.e., `adsl` and `adlb`.
- Step 2: Define the analysis dataset. In this example, we define two analysis datasets. The first dataset is the efficacy population (`gluc`). The second dataset is the collection of all records post baseline and on or before week 24 (`ana`).
- Step 3: Impute the missing values. In our example, we name the `ana` dataset after imputation as `ana_locf`.
- Step 4: Summarize `gluc`, i.e., calculate the mean and standard derivation, and then format it into an RTF table.
- Step 5: Summarize `ana_locf`, i.e., calculate the pairwise comparison by ANCOVA model, and then format it into an RTF table.
- Step 6: Rowly bind the output in Step 4 and Step 5.
